Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-08 manuscript.tex
1

Sensory data is most often represented by continuous values, such as the energy produced by a flow of photons which hits the different photoreceptors of the retina. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate of cortical cells in awake monkeys is highly irregular~\citep{softky_highly_1993}, which makes it inconsistent with temporal integration of firing rate. Remarkably, it has been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible, where the first spike is aligned, but the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{mainen_reliability_1995} (see Figure~\ref{fig:mainen}). However, if that same neuron is now driven by a \emph{frozen} noise, that is, a highly dynamic signal which is repeatedly replayed from trial to trial, then the output spikes are highly reproducible (for a review, see\citep{ermentrout_reliability_2008}). This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.'' 

@@

% retina
At the level of the dynamic processing of visual information, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work in the retina~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina  can be extended to the visual cortex.

% %----------------------------%
% \begin{figure}
% \centering
% \note{redo a figure similar to the thorpe code}%\includegraphics{images/roc.jpg}
% \caption{Latency coding. An input analog profile is encoded in latencies: the higher the contrast, the shorter the latency. In this example, one generates at most one spike per neuron.
% }\label{fig:roc}
% \end{figure}
% %----------------------------%
In fact, similar results have been demonstrated through neurophysiological recordings in the primary visual cortex and show that different levels of visual activity will induce different levels of neuronal discharge latency in the primary visual area~\citep{celebrini_dynamics_1993}. First-spike latency codes are a feasible mechanism for information transfer, even when biologically plausible estimates of stimulus onset are considered, for instance for sound localization~\citep{chase_first-spike_2007}.  Note also that timing is not entirely sensory or internal, but can be used as a general coding principle. In~\citep{safaie_turning_2020} for instance, they found that ``timing accuracy was improved when the environment afforded cues that rats can incorporate into motor routines. Timing, at least in animals, may thus be fundamentally embodied and situated.''  Many models have used these properties in temporal coding to build fast image categorization networks~\citep{gautrais_rate_1998,delorme_spikenet_1999, perrinet_coding_2004}. These models take the form of artificial spiking neural networks (SNNs) and have been able to demonstrate their practical applications for image categorization~\citep{delorme_ultra-rapid_2000}. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron by using the rank of neurons' activation~\citep{delorme_spikenet_1999,bonilla_analyzing_2022}. Another type of SNN using precise spike timing attempts to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been extended in novel gradient-based methods. The surrogate gradient method is now widely used in methods that attempt to transfer performance from CNNs to SNNs~\citep{zenke_remarkable_2021}. This type of modeling often uses the classical architecture of image categorization developed in deep learning, while adapting it to the specificity of the event-based representation~\citep{goltz_fast_2021}. For instance,~\citep{kheradpisheh_stdp-based_2018} implements a STDP-based spiking deep convolutional neural networks for object recognition or~\citep{tavanaei_representation_2018} develops a form of spike-based, competitive learning applied for unsupervised learning.

However, the performance of SNNs is still lagging compared to that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open. Achieving to efficiently implement this type of modeling is extremely important with respect to the development of a new generation of cameras called Silicon Retinas which represents the visual information as events, similarly to the representations described above and instead of using a basic frame-based representation~\citep{lichtsteiner_128-128_2008} (for a review, see~\citep{rasetto_challenges_2022}, this will be developped in Section~\ref{sec:neuromorphic}). % This work has been extended to include unsupervised learning capabilities, and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{grimaldi_homeostatic_2021,grimaldi_robust_2022}. 
To alleviate this gap in computational neuroscience, we will further review the role of  precise spike timing in neural assemblies from a neuroscientific perspective in the next section.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Role of precise spike timing in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One first hypothesis: synchronous firing in cell assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code that we have already explored at the cellular level~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that this could take the form of ``synfire chains,'' that is, the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.

>>>>

Sensory data is most often represented by continuous values, such as the energy produced by a flow of photons which hits the different photoreceptors of the retina. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate of cortical cells in awake monkeys is highly irregular~\citep{softky_highly_1993}, which makes it at first sight inconsistent with the temporal integration of firing rate. Remarkably, it has been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible: whike the first spike are aligned, the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{mainen_reliability_1995} (see Figure~\ref{fig:mainen}). However, if that same neuron is now driven by a \emph{frozen} noise, that is, a highly dynamic signal which is repeatedly replayed from trial to trial, then the output spikes are highly reproducible (for a review, see~\citep{ermentrout_reliability_2008}). This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.'' 

@@

% retina
How may information about the sensory world be encoded in the spiking pattern?  At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.

% %----------------------------%
% \begin{figure}
% \centering
% \note{redo a figure similar to the thorpe code}%\includegraphics{images/roc.jpg}
% \caption{Latency coding. An input analog profile is encoded in latencies: the higher the contrast, the shorter the latency. In this example, one generates at most one spike per neuron.
% }\label{fig:roc}
% \end{figure}
% %----------------------------%
In fact, similar results have been demonstrated through neurophysiological recordings in the primary visual cortex and show that different levels of visual activity will induce different levels of neuronal discharge latency in the primary visual area~\citep{celebrini_dynamics_1993}. First-spike latency codes are a feasible mechanism for information transfer, even when biologically plausible estimates of stimulus onset are considered, for instance for sound localization~\citep{chase_first-spike_2007}.  Note also that timing is not entirely sensory or internal, but can be used as a general coding principle. In~\citep{safaie_turning_2020} for instance, they found that ``timing accuracy was improved when the environment afforded cues that rats can incorporate into motor routines. Timing, at least in animals, may thus be fundamentally embodied and situated.''  Many models have used these properties in temporal coding to build fast image categorization networks~\citep{gautrais_rate_1998,delorme_spikenet_1999, perrinet_coding_2004}. These models take the form of artificial spiking neural networks (SNNs) and have been able to demonstrate their practical applications for image categorization~\citep{delorme_ultra-rapid_2000}. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron by using the rank of neurons' activation~\citep{delorme_spikenet_1999,bonilla_analyzing_2022}. Another class of artificial SNNs use precise spike timing as a metric in order to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been extended in novel gradient-based methods. The subsequent surrogate gradient method is now widely used in methods that attempt to transfer performance from analog (CNNs) to spike-based (SNNs) architectures~\citep{zenke_remarkable_2021}. This type of modeling often uses the classical task of categorizing images developed in deep learning, while adapting it to the specificity of the event-based representation~\citep{goltz_fast_2021}. For instance,~\citep{kheradpisheh_stdp-based_2018} implements a STDP-based spiking deep convolutional neural networks for object recognition or~\citep{tavanaei_representation_2018} develops a form of spike-based, competitive learning applied for unsupervised learning.

However, the performance of SNNs is still lagging compared to that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open. Achieving to efficiently implement this type of modeling is extremely important with respect to the development of a new generation of cameras called Silicon Retinas. These represent visual information as events, similarly to the biologically-inspired representation described above and in contrast with a basic frame-based representation~\citep{lichtsteiner_128-128_2008} (for a review, see~\citep{rasetto_challenges_2022}, this will be developped in Section~\ref{sec:neuromorphic}). % This work has been extended to include unsupervised learning capabilities, and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{grimaldi_homeostatic_2021,grimaldi_robust_2022}. 
To alleviate this gap in computational neuroscience, we will further review the role of  precise spike timing in neural assemblies from a neuroscientific perspective in the next section.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Role of precise spike timing in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One first hypothesis: synchronous firing in cell assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code that we have already explored at the cellular level~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that this could take the form of ``synfire chains,'' that is, the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.
